A Flux Scale for Southern Hemisphere 21cm EoR Experiments

### 2013-05-22 ms.tex
Updated description of source selection vis beam error.

\begin{abstract}
We present a catalog of broad-band spectral measurements from the 
Donald C. Backer Precision Array for Probing
% XXX ARP: was irina "first"?  and maybe "catalog"
the Epoch of Reionization (PAPER) in South Africa observed in July and
September of 2011.  In order to reduce the impact of beam calibration, which proves to
be a difficult endeavor for transit telescopes such as PAPER, on the determination
of source spectra, we have focused on calibrating sources in a narrow declination range.
Since each source follows a nearly identical path through the primary beam, this
restriction allows beam calibration to be nearly eliminated as a source of error,
yielding a dramatic improvement in the accuracy of source spectra measured in the 
100--200-MHz band that is receiving renewed attention by experiments seeking to
measure 21cm emission from the Epoch of Reionization (EoR).
Based on the variation in our data and the estimated error in the absolute
flux scale we bootstrap from multiple bright sources, we estimate PAPER measurements
of sources above XXX Jy to be accurate to 3.2\%.
% DCJ 3.2% is just for Pictor A, and certainly not all PAPER measurements of sources above XXX Jy
When combined with catalog data at other frequencies, these data constrain parameters of
a power-law model for flux density with an uncertainty of 2.4\%.
% XXX ARP: this must again be dependent on source strength and location.
This accuracy is limited by the uncertainty in the catalog measurements we use
to estimate an absolute flux scale, and represents 
an order of magnitude improvement over previous measurements in this band.  
Comparing with prior measurements, 45 out
of 58 sources observed are found to confirm and refine a power-law model for flux density.
This includes Pictor A, which provides a key flux reference for PAPER's EoR
power spectrum analysis.
\end{abstract}

@@

More recent surveys include narrow-band surveys by the GMRT and Mauritius, a
deep survey of the region near Hydra A by the 32 antenna MWA prototype
\cite{Williams:2012p8768} and a wide field survey by PAPER, also with 32
elements \cite{Jacobs:2011p8438}. Several sub-channels were provided in the
Williams catalog, though with 60-80\% error bars ---large compared to the 30\%
uncertainty on their wide band measurements.  The latter cover the band and
spatial scales relevant to EoR measurements but are limited by the accuracy of
the primary beam \citep{Jacobs:2013p9713}.

@@

This paper reports a set of flux measurements where a primary beam
model has not been used to estimate the flux, but instead are directly
calibrated to a single calibrator making the track through the beam.

EoR measurements in the southern sky have focused on the coldest regions where
galactic foregrounds are minimal. With the majority of possible observing time falling on 
RA=4h,Dec=-30. The brightest and least-resolved calibrator in this region is Pictor A 
(5h19m49.1 -45d46m45.0). Pictor A is a nearby FR-II type radio galaxy 
 similar to Cygnus A.  At $\sim$400Jy Pictor is bright and sufficiently distant from other 
bright sources to make it eminently suitable as both a phase and flux calibrator. Its apparent
size of $\sim$8\arcmin is smaller than the scales being probed by current EoR instruments , 
making resolution effects, making it suitable for precision calibration.  However, 
like most other sources, precise flux measurements in the EoR band are not available.
The current best EoR band measurement
is uncertain to 12\% and appears to imply spectral flattening in the EoR band
\citep{Perley:1997p9312}.

@@

\section{Approach}
\label{sec:approach}
Given the large uncertainties in the literature, and the necessity for PAPER power-spectrum analysis
to use Pictor A for flux calibration, we felt it prudent to examine Pictor A directly.
In this section, we describe how we use PAPER in its imaging configuration
to measure Pictor A and a selection of known, bright sources in a 5\arcdeg strip centered
on -45\arcdeg.

 In \citet{Jacobs:2013p9837} several limitations were identified that
exasperated the primary beam uncertainty.  First, in imaging each source flux is measured at just a few 
points in the primary beam. As errors tend to vary across the beam, the uncertainty will tend to vary between sources.
Here we have averaged over all possible positions making the uncertainty a function of the net uncertainty in the beam,
and stabilizing the source-to-source variations in error bars. The second limitation, identified in \citet{Williams:2012p8768},
was the increase of uncertainty towards the edge of the beam. To minimize our sensitivity to both beam and noise
uncertainty, we weight the source track by an additional factor of the primary beam model.

As PAPER is a drift scan instrument, each declination $\delta$ describes a distinct cut through the primary beam. These cuts
provide a detailed measurement of the primary beam relative to the peak of the trace, however,
a model $g(\delta)$ is still required to calibrate between declinations. This model will be effective over a range of declinations, but will diverge at
some rate depending on the accuracy with which the model describes the slope.  As an estimate of the extrapolated 
beam amplitude we fit the slope of our Pictor A source track and compute the amplitude extrapolated over our fiducial
stripe width of 5\arcdeg.

\[
B = dB/dt dt/\theta \Delta \theta
\]

Comparing this with the simulation in Figure \ref{fig:beam_extrapolation}, we find that the model predicts the 5\arcdeg extrapolation to better
than 2\% even into the more uncertain areas far from zenith. The RMS difference for the central 15\arcdeg of the track
is 1.5\%. From this we conclude that, as long as we constrain our extrapolations to $\pm5$\arcdeg around Pictor A,
we will have negligible beam area.

\begin{figure*}
\includegraphics[width=.9\textwidth]{plots/beam_slope_compare.png}
\caption{The beam is well along a declination cut but beam formed source tracks. Here we measure the beam, 
as extrapolated by fitting a slope (dotted), and compare with the beam model (solid). The difference (circles) is never larger
than $\sim$1\%.  Thus we conclude that the beam slope is locally well modeled and our beam model is suitable 
for extrapolation over $\sim$5\arcdeg scales.
  \label{fig:beam_extrapolation}}
\end{figure*}

@@

\section{Results and Discussion}
\label{sec:fits}

%0008-421#
%1243-412#
%1247-401#
%1315-460
%TODO round up the source counts

In the 59 sources measured, four demonstrated poor convergence fitting a single
spectral index to both the PAPER data and prior catalog data. In three
cases\footnote{1243-412,1247-401,1315-460} %TODO update these sources are in
the Centaurus A/galactic plane region. Though the beam forming method provides
relatively high dynamic range of $\sim$100:1 it is unable to distinguish
between these sources and the bright diffuse Centaurus and galactic structure.
The last source, 0008-421, was not detected by previous PAPER observations and
exhibits flattening at higher frequencies, most likely due to synchrotron
self-absorption \citep{Jacobs:2011p8438} .  The remaining 94\% MCMC chains
arrived at a stable posterior distribution. The spectral index parameters for
these chains are given in table \ref{tab:fits}.

>>>>

\begin{abstract}
We present a catalog of broad-band spectral measurements from the 
Donald C. Backer Precision Array for Probing
the Epoch of Reionization (PAPER) in South Africa observed in July and
September of 2011.  In order to reduce the impact of beam calibration, which proves to
be a difficult endeavor for transit telescopes such as PAPER, on the determination
of source spectra, we have focused on calibrating sources in a narrow declination range.
Since each source follows a nearly identical path through the primary beam, this
restriction allows beam calibration to be greatly reduced as a source of error,
yielding a dramatic improvement in the accuracy of source spectra measured in the 
100--200-MHz band which is receiving renewed attention by experiments seeking to
measure 21cm emission from the Epoch of Reionization (EoR). Using beam forming
and fringe-rate filtering we measure spectra for Pictor A, which provides a key flux reference for PAPER's EoR
power spectrum analysis, 
as well as 60 other sources at nearby declinations. Extrapolating from higher frequency
catalogs we set the flux scale in a monte-carlo fit across multiple sources. This fit 
propagates uncertainty from both catalog and measurement errors. Fitting new spectral 
models to both catalog and the new PAPER data we find a Pictor A flux model accurate to
2.4\%.
This accuracy is limited by the uncertainty in the catalog measurements we use
to estimate an absolute flux scale, and represents 
an order of magnitude improvement over previous measurements in this band.  
Comparing with prior measurements, 45 out
of 58 sources observed are found to confirm and refine a power-law model for flux density.

\end{abstract}

@@

More recent surveys include narrow-band surveys by the GMRT and Mauritius, a
deep survey of the region near Hydra A by the 32 antenna MWA prototype
\cite{Williams:2012p8768} and a wide field survey by PAPER, also with 32
elements \cite{Jacobs:2011p8438}. Several sub-channels were provided in the
Williams catalog, though with 60-80\% error bars ---large compared to the 30\%
uncertainty on their wide band measurements.  The latter cover the band and
spatial scales relevant to EoR measurements but are limited by the accuracy of
the primary beam \citep{Jacobs:2013p9713}. The accuracy of the absolute flux calibrators in the southern hemisphere
is similarly limited.

The response of the primary beam is of critical importance to EoR measurements.
Differences between the polarization responses cause leakage of polarized
signals into the total intensity measurement possibly corrupting the 
EoR power spectrum \cite{Moore:2013p9621}.  The primary beam 
shape is also critical to measuring and subtracting foregrounds 
\cite{Bernardi:2012p9020} a process which, to be affective, must be done to better than 1\%
precision for the brightest sources \citet{Liu:2009p4762,Bowman:2009p7816} .

@@

In this paper we provide a set of detailed total flux measurements aimed at providing
an absolute flux calibration for southern hemisphere EoR studies as well as a
small set of verification fluxes suitable for constraining future primary beam studies.

\subsection{The sources and instrument}
EoR measurements in the southern sky have focused on the coldest regions where
galactic foregrounds are minimal. With the majority of possible observing time falling on 
RA=4h,Dec=-30. The brightest and least-resolved calibrator in this region is Pictor A 
(5h19m49.1 -45d46m45.0). Pictor A is a nearby FR-II type radio galaxy 
 similar to Cygnus A.  At $\sim$400Jy Pictor is bright and sufficiently distant from other 
bright sources to make it eminently suitable as both a phase and flux calibrator. Its apparent
size of $\sim$8\arcmin is smaller than the scales being probed by current EoR instruments , 
making resolution effects, making it suitable for precision calibration.  However, 
like most other sources, precise flux measurements in the EoR band are not available.
The current best EoR band measurement
is uncertain to 12\% and appears to imply spectral flattening in the EoR band
\citep{Perley:1997p9312}.

@@

\section{Approach}
\label{sec:approach}
In this section, we describe how we use PAPER in its imaging configuration
to measure Pictor A and a selection of known, bright sources suitable for confirming the accuracy of the measurement
and extrapolated flux scale.

 In \citet{Jacobs:2013p9837} several limitations were identified that
exasperated the primary beam uncertainty.  First, under most imaging schemes, each source flux is measured at just a few 
points in the primary beam. As errors tend to vary across the beam, the uncertainty will tend to vary between sources.
Here we have averaged over long tracks through the primary beam. Using the drift scan-beamforming technique, each 
source is measured hundreds of times at a variety of different pointings. The variation between these measurements is then
used to calculate measurement uncertainty. Second, the flux scale was found to vary substantially over the 
sky, due to the dual uncertainties of primary beam response and prior catalog. As a hedge against beam uncertainty 
we limit our selection to source with similar primary beam responses. A third limitation, identified in \citet{Williams:2012p8768},
was the increase of uncertainty towards the edge of the beam. To minimize our sensitivity to both beam and noise
uncertainty, we weight the source track by an additional factor of the primary beam model.

\subsection{Selection of confirmation sources}
As PAPER is a drift scan instrument, each declination describes a distinct path through the primary beam. These paths
provide a detailed sample of the primary beam relative to the peak of the trace, however,
a model is still required to calibrate between declinations and the accuracy of the trace is sometimes contaminated by
sidelobes. In section XXX we will compare the beam response as measured by the PicA trace with the model 
used to estimate flux. We will see that, on average, the primary beam is not the limiting factor in estimating the flux and that
limiting ourselves to a declination range of 10\arcdeg around Pictor A for flux verification is a good compromise.
%
%
%
%This model will be effective over a range of declinations, but will diverge at
%some rate depending on the accuracy with which the model describes the slope.  As an estimate of the extrapolated 
%beam amplitude we fit the slope of our Pictor A source track and compute the amplitude extrapolated over our fiducial
%stripe width of 5\arcdeg.
%
%\[
%B = dB/dt dt/\theta \Delta \theta
%\]
%
%Comparing this with the simulation in Figure \ref{fig:beam_extrapolation}, we find that the model predicts the 5\arcdeg extrapolation to better
%than 2\% even into the more uncertain areas far from zenith. The RMS difference for the central 15\arcdeg of the track
%is 1.5\%. From this we conclude that, as long as we constrain our extrapolations to $\pm5$\arcdeg around Pictor A,
%we will have negligible beam area. 
%
%\begin{figure*}
%\includegraphics[width=.9\textwidth]{plots/beam_slope_compare.png}
%\caption{The beam is well along a declination cut but beam formed source tracks. Here we measure the beam, 
%as extrapolated by fitting a slope (dotted), and compare with the beam model (solid). The difference (circles) is never larger
%than $\sim$1\%.  Thus we conclude that the beam slope is locally well modeled and our beam model is suitable 
%for extrapolation over $\sim$5\arcdeg scales.
%  \label{fig:beam_extrapolation}}
%\end{figure*}

@@

\section{Results and Discussion}
\label{sec:fits}
In the 59 sources measured, four demonstrated poor convergence fitting a single
spectral index to both the PAPER data and prior catalog data. In three
cases\footnote{1243-412,1247-401,1315-460} %TODO update these sources are in
the Centaurus A/galactic plane region. Though the beam forming method provides
relatively high dynamic range of $\sim$100:1 it is unable to distinguish
between these sources and the bright diffuse Centaurus and galactic structure.
The last source, 0008-421, was not detected by previous PAPER observations and
exhibits flattening at higher frequencies, most likely due to synchrotron
self-absorption \citep{Jacobs:2011p8438} .  The remaining 94\% MCMC chains
arrived at a stable posterior distribution. The spectral index parameters for
these chains are given in table \ref{tab:fits}.
